Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev

### 2024-09-28 main.tex
Addressed many of Benjamin's feedback comments.

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add an
  automatically controlled steering motor to a consumer electric bicycle that
  has a stabilizing effect down to about 3~\si{\kph}. We hypothesize that a
  stabilized bicycle will reduce the probability of falling. To test the
  motor's possible assistance during falls, we apply varying magnitude external
  handlebar perturbations to twenty-six participants who rode on a treadmill
  with the balance assist system activated and deactivated. The probability of
  recovering from a handlebar perturbation significantly increases when the
  balance assist is activated at a travel speed of 6~\si{\kph}. This positive
  effect is most prominent at and around the individual riders' perturbation
  resistance threshold. We conclude that use of a balance assist system in real
  world bicycling would reduce the number of falls that occur near riders'
  control authority limits.
}

@@

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles are not self-stable and can be
challenging for the rider to balance. Low speed crashes may be reduced if the
bicycle was self-stable at these speeds by relieving the rider of some of their
necessary control activity. Bicycles can be mechanically modified to lower the
speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
has been known that applying a motor actuated steering torque proportional to
the vehicle's roll angular rate~\footnote{Roll angular rate is defined herein
as per the SAE J670 vehicle standards.} can stabilize a single-track vehicle
down to very low speeds~\citep{Ruijs1986}. If automatic control of steering can
stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks much like the natural high speed
self-stability already does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

In this paper, we test whether a steer motor controlled bicycle, that is stable
in a large low speed range, is beneficial in helping to prevent the rider from
falling. We test this by applying varying magnitude mechanical perturbations to
the handlebars while the rider is cycling on a treadmill. We assess the
difference in the rider's probability of falling with the balance assist system
on and off.

@@

If the steer torque is the sum of the (h)uman applied torque and the (m)otor
applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
\(\mathbf{B} = \begin{bmatrix} \vec{B}_\phi \quad \vec{B}_\delta
\end{bmatrix}\), and the motor torque is a proportional feedback controller
\(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then the human-controlled
plant takes the form:
%
\begin{align}
  \dot{\vec{x}} = \left(\mathbf{A} -
  \vec{B}_\delta
  \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right]
\right)
  \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
\end{align}

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified an electric bicycle (Grenoble C8 HMB, Royal Dutch Gazelle, Dieren,
The Netherlands) with a custom motor in the steering assembly capable of
applying up to 7~\si{\newton\meter} of torque between the head tube and steer
tube, see Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller
converts the commanded torque to applied torque. We measure the speed of the
rear wheel with an encoder (ABS Speed Sensor, Magura GmbH, Bad Urach, Germany)
and measure the body fixed roll rate of the bicycle with a MEMs rate gyroscope
(VR IMU BN0086, Sparkfun, Niwot, USA). The balance assist control algorithm is
implemented on a microprocessor (Teensy, PJRC, USA) and data from all sensors
is logged with a CAN bus (CanEdge2, CSS Electronics, Aabyhoej, Denmark) at at
least 100 Hz.

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll angular rate
feedback controller to stabilize the bicycle. In the speed range tested, the
commanded steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the
control law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use \(g=8\) (low) and \(g=10\) (high) to tune the gain magnitude during the
experiments.  Scaling the proportional feedback gain linearly with respect to
speed stabilizes the normally unstable weave mode of the bicycle down to
2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled with \(g=10\) (lower row)
    root locus of eigenvalue components (real: solid, imaginary: dashed)
    plotted versus speed for the bicycle without the inertial effect of a rigid
    rider (left) and with a rigid rider (right). Vertical dotted and
    dotted-dashed lines indicate the two speeds we perform experiments at. Grey
    shaded region is the linear stable speed range. Geometric and inertial
    parameters for these plots were estimated with the methods presented in
    \citet{Moore2012} and software packages BicycleParameters
    1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015} and are shown
    in Appendix~\ref{sec:physical-parameters}.}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar from Earth fixed
posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
four motors (EC-90 Flat, Maxon Group, Switzerland) working in cooperation,
Figure~\ref{fig:setup-diagram}. The four motors are programmed to apply a light
force at all times to keep the ropes taught and to track a commanded force
profile using a PID controller running on a microprocessor (Arduino Mega 2560,
Arduino, Italy). We measure the force applied by each motor at the handlebar
via four inline S-style load cells (Bosche GmbH, Damme, Germany). All were
rated for 250~\si{\newton} except the left rear load cell which was rated for
500~\si{\newton}. The commanded force profiles are designed to apply an
external pulsive torque to the front assembly (handlebars, forks, wheel) at
magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four motors are
arranged at the four corners of a 1~\si{\meter} wide by 2~\si{\meter} long
treadmill (Bonte Technology B.V., Zwolle, The Netherlands) that can reach
speeds up to 18~\si{\kilo\meter\per\hour}. The general design of this
perturbation system is described in detail in Van De Velde's MSc
thesis~\citep{vandeVelde2022a} and the physical arrangement is shown in
Figure~\ref{fig:participant-in-set-up}. Our modifications relative to Van De
Velde's design include simplifying the Bump'Em motor controller with an
inexpensive microcontroller and the use of a simpler non-active safety harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[height=65mm]{figures/participant-in-set-up.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations and the various system components.}
  \label{fig:experimental-setup}
\end{figure}

@@

The participants were divided into two groups. The first group of fifteen
participants performed the protocol at a belt speed of
10~\si{\kph}(2.8~\si{\mps}) with the controller gain value set to $g=8$ and the
second group of eleven particpants performed the protocol at a belt speed of
6~\si{\kph} (1.7~\si{\mps}) with the controller gain set to $g=10$. The
10~\si{\kph} group's experiments occurred two weeks before the 6~\si{\kph}
group.

@@

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure applying perturbations
above and below the initial perturbation threshold, while allowing small
progression of the perturbation threshold to accommodate learning effects. Five
possible perturbation forces are determined based on the initial perturbation
threshold estimate: the initial estimate itself, two perturbations lower than
the initial estimate, and two perturbations higher than the initial estimate.
The five possible perturbations are separated by 10~\si{\newton} steps. For
example, if the initial estimate of the perturbation threshold of a participant
is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
100~\si{\newton}. Which one of these five perturbations is chosen first is
determined at random. If the perturbation results in a fall, the estimate of
the perturbation threshold is decreased by 10~\si{\newton}, and vice versa if
the perturbation did not result in a fall. Five new possible perturbation
forces are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied. The goal of this adaptive staircase procedure is to
have participants fall for approximately 50\% of the time. After the first set
of 20 perturbations, we let the cyclist rest and then perform another 20
perturbations. We randomize whether the balance assist system is on or off
during the first or second set of 20 perturbations for each participant.

@@

Based on findings from measuring riders without balance assist in
\citet{Haitjema2023a}, we calculate several variables that we hypothesize may
influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}(F_r + F_l) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[(F_{rf} - F_{rr}) + (F_{lf}
  - F_{lr})\right] dt
  \textrm{.}
  \label{eq:angular-impulse}
\end{align}

where \(F_r\) and \(F_l\)  is the total force applied on the right and left
handlebar ends, respectively which are the sum of the rear and front load cell
readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
as \(l\) in Equation \ref{eq:angular-impulse}. We use angular impulse instead
of peak torque to normalize for the duration of the applied perturbation.

We record the order number of each perturbation \(j\), i.e. first, second,
third, \ldots, to measure how many perturbations a rider is subjected to before
the current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles to characterize the configuration of the
bicycle when perturbed. The gain setting on the balance assist controller
indicates if the assistance is off \(g=0\) or on at two different levels: low
\(g=8\) or high \(g=10\). A recovery from the perturbation is successful if the
person neither places their foot down onto the treadmill surface nor their
wheel of the bicycle exits the width of the treadmill belt. We record this as a
binary variable \(f\) for "fall". All measured variables are reported in
Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall Outcome'' and
    ``Perturbation Order Number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall Outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering, as recommended by \citet{Enders2007}, with clusters being an
individual participant. The clusters are chosen as all data from an individual
participant because we are only interested in the association between the state
of the balance assist system and the outcome of the perturbation. We expected
there to be a variation between participants in how well they are able to
resist the perturbation. However, this was not true. Cluster-mean centering
showed there to be no variation between participants. This fact allows us to
utilize a simple single-level logistic regression model, instead of a
multilevel model. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(g=10\) trials and one
for the 10~\si{\kph}, \(g=8\) trials and we evaluate our hypothesis for each
set of data. Statistics were computed with the lme4 version
1.1~\citep{Bates2015} package in R version 4.1.3~\citep{RCoreTeam2021}.

\section{Results}
%
The coefficient estimates for a single-level logistic regression for the data
from the 6~\si{\kph}, gain $g=10$ trials are shown in
Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
balance assist state are all statistically significant predictors with angular
impulse having a dominant effect. Larger angular impulse increases the
probability to fall and both enduring more perturbations or having the balance
assist on, decrease the probability to fall. The associated multiplicative
change in odds are shown in Table~\ref{tab:freq-coefs-6}.
%
\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates \(\alpha_k\) at
    6~\si{\kph} and gain $g=10$ along with the standard error SE, p-value
    \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
    confidence interval bounds.}
  \label{tab:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\alpha_k\) & SE & \(p\) & \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept \(\beta\)                           & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse \(L\)                         & 1.69 & 0.27 & 0.00\(^*\) & 5.40 & 3.18 & 9.16 \\
    Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
    Balance assist state \(c\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
    Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle \(\delta_0\)                      & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

>>>>

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add an
  automatically controlled steering motor to a consumer electric bicycle that
  has a stabilizing effect down to about 4~\si{\kph}. We hypothesize that a
  stabilized bicycle will reduce the probability of falling. To test the
  motor's possible assistance during falls, we apply varying magnitude external
  handlebar perturbations to twenty-six participants who rode on a treadmill
  with the balance assist system activated and deactivated. The probability of
  recovering from a handlebar perturbation significantly increases when the
  balance assist is activated at a travel speed of 6~\si{\kph}. This positive
  effect is most prominent at and around the individual riders' perturbation
  resistance threshold. We conclude that use of a balance assist system in real
  world bicycling would reduce the number of falls that occur near riders'
  control authority limits.
}

@@

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles are not self-stable and can be
challenging for the rider to balance. Low speed crashes may be reduced if the
bicycle was self-stable at these speeds by relieving the rider of some of their
necessary control activity. Bicycles can be mechanically modified to lower the
speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
has been known that applying a motor actuated steering torque proportional to
the vehicle's roll angular rate~\footnote{Roll angular rate is defined herein
as per the SAE J670 vehicle standards.} can stabilize a single-track vehicle
down to very low speeds~\citep{Ruijs1986}. If automatic control of steering can
stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks much like the natural high speed
self-stability already does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack first introduced in \citet{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

In this paper, we investigate whether an automatically controlled bicycle, that
is stable in a large low speed range for balance assist, is beneficial in
helping to prevent the rider from falling. We test this possible benefit by
applying varying magnitude mechanical perturbations to the handlebars while the
rider is cycling on a treadmill. We assess the difference in the rider's
probability of falling with the balance assist system on and off.

@@

If the steer torque is the sum of the (h)uman applied torque and the (m)otor
applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
\(\mathbf{B} = \begin{bmatrix} \vec{B}_\phi \quad \vec{B}_\delta
\end{bmatrix}\), and the motor torque is a proportional feedback controller
\(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then the human-controlled
plant takes the form:
%
\begin{align}
  \dot{\vec{x}} = \left(\mathbf{A} -
  \vec{B}_\delta
  \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right]
\right)
  \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
  \textrm{.}
\end{align}

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified an electric bicycle (Grenoble C8 HMB, Royal Dutch Gazelle, Dieren,
The Netherlands) with a custom motor in the steering assembly capable of
applying up to 7~\si{\newton\meter} of torque between the head tube and steer
tube, see Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller
converts the commanded torque to applied torque. We measure the speed of the
rear wheel with an encoder (ABS Speed Sensor, Magura GmbH, Bad Urach, Germany)
and measure the body fixed roll rate of the bicycle with a MEMs rate gyroscope
(VR IMU BN0086, Sparkfun, Niwot, USA). The balance assist control algorithm is
implemented on a microprocessor (Teensy, PJRC, USA) and data from all sensors
is logged with a CAN bus (CanEdge2, CSS Electronics, Aabyhoej, Denmark) at
least 100 Hz.

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll angular rate
feedback controller to stabilize the bicycle. In the speed range tested, the
commanded steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the
control law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use the factor \(g=8\) (low) and \(g=10\) (high) to tune the gain magnitude
during the experiments. Scaling the proportional feedback gain linearly with
respect to speed stabilizes the normally unstable weave mode of the bicycle
down to 2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled with \(g=10\) (lower row)
    root locus of eigenvalue components (real: solid, imaginary: dashed)
    plotted versus speed for the bicycle without the inertial effect of a rigid
    rider (left) and with a rigid rider (right). Vertical dotted and
    dotted-dashed lines indicate the two speeds we perform experiments at. Grey
    shaded region is the linear stable speed range. Geometric and inertial
    parameters for these plots were estimated with the methods presented in
    \citet{Moore2012} and software packages BicycleParameters
    1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015} and are shown
    in Appendix~\ref{sec:physical-parameters}.}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar from earth-fixed
posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
four motors (EC-90 Flat, Maxon Group, Switzerland) working in cooperation,
Figure~\ref{fig:setup-diagram}. The four motors are programmed to apply a light
force at all times to keep the ropes taught and to track a commanded force
profile using a PID controller running on a microprocessor (Arduino Mega 2560,
Arduino, Italy). We measure the force applied by each motor at the handlebar
via four inline S-style load cells (Bosche GmbH, Damme, Germany). All were
rated for 250~\si{\newton} except the left rear load cell which was rated for
500~\si{\newton}. The commanded force profiles are designed to apply an
external pulsive torque to the front frame at magnitudes varying from
\SIrange{16}{160}{\newton\meter}. The four motors are arranged at the four
corners of a 1~\si{\meter} wide by 2~\si{\meter} long treadmill (Bonte
Technology B.V., Zwolle, The Netherlands) that can reach speeds up to
18~\si{\kilo\meter\per\hour}. The general design of this perturbation system is
described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
Our modifications relative to Van De Velde's design include simplifying the
Bump'Em motor controller with an inexpensive microcontroller and the use of a
simpler non-active safety harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[height=65mm]{figures/participant-in-set-up.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations and the various system components.}
  \label{fig:experimental-setup}
\end{figure}

@@

The participants were divided into two groups. The first group of fifteen
participants performed the protocol at a belt speed of
10~\si{\kph}(2.8~\si{\mps}) with the controller factor set to $g=8$ and two
weeks later the second group of eleven particpants performed the protocol at a
belt speed of 6~\si{\kph} (1.7~\si{\mps}) with the controller factor set to
$g=10$.

@@

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure applying perturbations
above and below the initial perturbation threshold, while allowing small
progression of the perturbation threshold to accommodate learning effects. Five
possible perturbation forces are determined based on the initial perturbation
threshold estimate: the initial estimate itself, two perturbations lower than
the initial estimate, and two perturbations higher than the initial estimate.
The five possible perturbations are separated by 10~\si{\newton} steps. For
example, if the initial estimate of the perturbation threshold of a participant
is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
100~\si{\newton}. Which one of these five perturbations is chosen first is
determined at random. If the perturbation results in a fall, the estimate of
the perturbation threshold is decreased by 10~\si{\newton}, and vice versa if
the perturbation did not result in a fall. Five new possible perturbation
forces are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied. The goal of this adaptive staircase procedure is to
have participants fall for approximately 50\% of the time. After the first set
of 20 perturbations, we let the cyclist rest and then perform another 20
perturbations. We randomize whether the balance assist system is on or off
during the first or second set of 20 perturbations for each participant and the
participants were not told the order.

@@

Based on findings from measuring riders without balance assist in
\citet{Haitjema2023a}, we calculate several variables that we hypothesize may
influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}(F_r + F_l) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[(F_{rf} - F_{rr}) + (F_{lf}
  - F_{lr})\right] dt
  \label{eq:angular-impulse}
\end{align}
%
where \(F_r\) and \(F_l\)  is the total force applied on the right and left
handlebar ends, respectively which are the sum of the rear and front load cell
readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
as \(l\) in Equation \ref{eq:angular-impulse}. We use angular impulse instead
of peak torque to normalize for the duration of the applied perturbation.

We record the order number of each perturbation \(j\), i.e. first, second,
third, \ldots, to measure how many perturbations a rider is subjected to before
the current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles to characterize the configuration of the
bicycle when perturbed. The gain factor setting on the balance assist controller
indicates if the assistance is off \(g=0\) or on at two different levels: low
\(g=8\) or high \(g=10\). A recovery from the perturbation is successful if the
person neither places their foot down onto the treadmill surface nor their
wheel of the bicycle exits the width of the treadmill belt. We record this as a
binary variable \(f\) for ``fall''. All measured variables are reported in
Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall Outcome'' and
    ``Perturbation Order Number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall Outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering, as recommended by \citet{Enders2007}, with clusters being an
individual participant. The clusters are chosen as all data from an individual
participant because we are only interested in the association between the state
of the balance assist system and the outcome of the perturbation. We expected
there to be a variation between participants in how well they are able to
resist the perturbation; however, this was not true. Cluster-mean centering
showed there to be no variation between participants. This fact allows us to
utilize a simple single-level logistic regression model, instead of a
multilevel model. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(g=10\) trials and one
for the 10~\si{\kph}, \(g=8\) trials and we evaluate our hypothesis for each
set of data. Statistics were computed with the lme4 version
1.1~\citep{Bates2015} package in R version 4.1.3~\citep{RCoreTeam2021}.

\section{Results}
%
The coefficient estimates for a single-level logistic regression for the data
from the 6~\si{\kph}, gain factor $g=10$ trials are shown in
Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
balance assist state are all statistically significant predictors with angular
impulse having a dominant effect. Larger angular impulse increases the
probability to fall and both enduring more perturbations or having the balance
assist on, decrease the probability to fall. The associated multiplicative
change in odds are shown in Table~\ref{tab:freq-coefs-6}.
%
\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates \(\alpha_k\) at
    6~\si{\kph} and gain factor $g=10$ along with the standard error SE, p-value
    \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
    confidence interval bounds.}
  \label{tab:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\alpha_k\) & SE & \(p\) & \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept \(\beta\)                           & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse \(L\)                         & 1.69 & 0.27 & 0.00\(^*\) & 5.40 & 3.18 & 9.16 \\
    Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
    Balance assist state \(c\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
    Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle \(\delta_0\)                      & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

The coefficient estimates for a single-level logistic regression at
10~\si{\kph} with gain factor $g=8$ are shown in Table \ref{tab:freq-coefs-10}. The
angular impulse and perturbation order are statistically significant predictors
with angular impulse being the dominant effect. Larger angular impulse
increases the probability to fall and enduring more perturbations decreases the
probability to fall. Unlike the 6~\si{\kph} trials, the balance assist state is
not a significant predictor but the effect would be a reduction in the
probability to fall if it were. At both speeds, the angular impulse has about
twice the magnitude in effect as the perturbation order.
%
\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates \(\alpha_k\) at
    10~\si{\kph} and gain factor $g=8$ along with the standard error SE, p-value
    \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
    confidence interval bounds.}
  \label{tab:freq-coefs-10}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\alpha_k\) & SE & \(p\) & \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept                & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
    Angular impulse \(L\)    & 2.39  & 0.29 & 0.00\(^*\) & 10.92 & 6.23 & 19.13 \\
    Perturbation order \(j\) & -1.16 & 0.21 & 0.00\(^*\)  & 0.31 & 0.21 & 0.48 \\
    Balance assist on \(c\)  & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
    Roll angle \(\phi_0\)    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
    Steer angle \(\delta_0\) & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
    Balance assist state $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
    Balance assist state $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
    Balance assist state $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
    Balance assist state $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
    \bottomrule
  \end{tabular}
\end{table}
